---
title: "ASSIGNMENT"
author: "author"
date: "12/2/2021"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 1
1. A hormone level in humans is believed to be a chi-square distribution. A patient data
sample indicates that the variance of the hormone level is 4. What is the standard
deviation of the median hormone level in humans? In what proportion of humans is the
hormone level less than 1? In what proportion of humans is the hormone level between
1.5 and 3?  

### Answer a:  
```{r}
variance <- 4
std.dev <- 2

##std.dev = 2* degrees of freedom 
degres_of_freedom = std.dev/2
  
pchisq(1,degres_of_freedom)
```
The standard deviation of the median hormone level is 2.
68% of the humans have a hormone level less than 1.  

### Answer b:  
```{r}
# P(1.5 < x < 3) = P(x < 3) - P(x < 1.5) 
pchisq(3,2) - pchisq(1.5,2)
```

25% of the humans have a hormone level between
1.5 and 3.    

<br>

### Question 2
2. Approximate the distribution of the hormone-level in Question 1 above with a log-normal
distribution. What are your new answers to the 3 questions asked in Question 1?  

### Answer:
```{r}
dlnorm(1,meanlog=0, sdlog=1)

plnorm(3,0,1) - plnorm(1.5,0,1)
```
40% of the humans have a hormone level less than 1 while 21% of the humans have a hormone level between
1.5 and 3.  

<br>

### Question 3  
3. The data below is on vitamin D levels for several patients. A physician wants to know the
standard deviations of the mean and median vitamin D levels. What would you tell her?
17 25 17 33 24 15 18 19 14 23 19 23 25 27 21 26 20  

### Answer:
```{r}
VitaminData = c(17,25,17,33,24,15,18,19,14,23,19,23,25,27,21,26,20)
sd(VitaminData)
sv  <- sd(VitaminData)/sqrt(length(VitaminData)) 
sv
```
The mean and median vitamin D levels are not equal.  

<br> 

### Question 4
4. Find the “best” Linear Regression model that explains Y in the following dataset:
LassoRR_sysm590_4K5K70pct.csv. That is, recognize, but ignore multicollinearity and
do the usual things to find a Linear Regression model. 

```{r}
library(readr)
lassoData <- read_csv("LassoRR_sysm590_4K5K70pct.csv")
##colnames(lassoData)

cor(lassoData)
##Evidence of multicoliniarity given the high correlation among the x variables 
regresults <-  lm(Y ~ . ,lassoData)
summary(regresults)
```

From the correlation numbers above, we can see that there is existence of multicolinearity. The best linear regression model that explains Y can be seen at X32 intercept since it has the least significant value.  

<br>

### Question 5
5. Now, without ignoring multicollinearity for the dataset given you in Question 3, find the
“best” model using Ridge Regression.

```{r}
##Ridge regression
#Automatically does the ridge regression selecting the best lambda parameter
library(ridge)
mod<-linearRidge(Y ~.,data=lassoData,lambda="automatic"); 
summary(mod)

```


